### 2024-09-27 teaching/math240f24/draft.tex
sep 26

\vskip \medskipamount \hbox to\hsize {\relax \unhcopy \strutbox \cont
{{\tenssbx I. Foundations \kern 3pt}}{2}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 1.\kern .5em Set theory }{3}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 2.\kern .5em Propositional logic}{10}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 3.\kern .5em Predicate logic}{17}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 4.\kern .5em Proofs}{21}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 5.\kern .5em Functions}{28}}
%\smallskip
%\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 6.\kern .5em Cardinality}{33}}
%\smallskip
%\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 7.\kern .5em Relations}{37}}

@@

We define $g : F\to E$ by
$$g(A) = \cases{A\setminus\{x\}, & if $x\in A$;\cr A\cup \{x\}, & if $x\notin A$.\cr}$$
As before, $\bigl|g(A)\bigr|$ is even, since $A$ is assumed to be a member of $F$ now. Now
for any $A\in E$,
$$\eqalign{
g\bigl(f(A)\bigr) &= \cases{g\bigl(A\setminus\{x\}\bigr), & if $x\in A$;\cr
g\bigl(A\cup \{x\}\bigr), & if $x\notin A$\cr} \cr
&= \cases{\bigl(A\setminus\{x\}\bigr)\cup \{x\}, & if $x\in A$;\cr
\bigl(A\cup \{x\}\bigr)\setminus \{x\} , & if $x\notin A$\cr} \cr
&= \cases{A, & if $x\in A$;\cr
A, & if $x\notin A$\cr} \cr
&= A.\cr
}$$
The proof that $f\bigl(g(A)\bigr) = A$ is similar. Thus $g$ is the inverse of $f$.\slug
\bye

@@

\datestamp{9}{}
Earlier, we defined the cardinality of a set to be the number of elements it contains. What, then,
is the cardinality of $\NN$? How about $\RR$? You might say $\infty$, but this is not a number
(at least, it's not an element of $\NN$, the way all cardinalities of finite sets are).
So perhaps we should amend our question to the following: {\sl When do infinite sets have the same size?}
Our experience with functions leads us to the answer: {\sl When there exists a bijection between
them}. We shall say that $A$ and $B$ are {\it equipotent} (or {\it equinumerous}, or {\it have
the same cardinality}) if there exists a bijection between $A$ and $B$. In this case
we write $|A| = |B|$.

@@

So we can find subset of $\NN$ equipotent with it. It turns out we can also find supersets of $\NN$ with
the same property.
\proclaim Theorem \advthm. We have $|\NN| = |\ZZ|$.

@@

To show that $f$ is injective, there are four cases. Note first that if $n$ is even, then $f(n)\ge 0$,
and if $n$ is odd, then $f(n) < 0$. So if $f(m) = f(n)$ for some $m,n\in \NN$, then $f(m)$ and $f(n)$ must
either both be negative, or both be nonnegative. Either way, $m$ and $n$
are either both even or they are both odd. If $m$ and $n$ are both even, then from $f(m) = f(n)$ we derive
$${m\over 2} = {n\over 2},$$
whence multiplying by $2$ on both sides we see that $m=n$. If $m$ and $n$ are both odd, then
$$-{m+1\over 2} = -{n+1\over 2},$$
so, multiplying by $-2$ and subtracting $1$ from both sides we have $m=n$ in this case as well.

@@

\proof
Enumerate $A = \{a_0, a_1, \ldots\}$ and $B = \{b_0, b_1, \ldots\}$. Now given
$(a_i, b_j)\in A\times B$ for some $i,j\in \NN$, we can let
$$f(a_i, b_j) = 2^i3^j.$$
We show that this defines an injective function. Suppose that $f(a_i, b_j) = f(a_{i'}, f(b_{j'}).$
Then $2^i3^j = 2^{i'} 3^{j'}$, and by the uniqueness of prime factorisations in the Fundamental
Theorem of Arithmetic, we see that $i = i'$ and $j=j'$.

@@

Let $f : \NN\to A$. We shall show that $f$ is not surjective. For all $m,n\in \NN$, let $a_{m,n}$
be the $n$th bit of $f(m)$. Consider the infinite binary string
$$s = (1-a_{0,0},1-a_{1,1}, 1-a_{2,2}, 1-a_{3,3}, \ldots).$$
This string $s$ cannot equal $f(m)$ for any $m\in \NN$, since given an arbitrary $m$,
the $m$th bit of $f(m)$ is $a_{m,m}$, whereas the $m$th bit of $s$ is $1-a{m,m}$. Hence there is some
$s\in A$ such that $f(m)\ne s$ for all $m\in \NN$. So $f$ is not surjective.

@@

The set of all infinite binary strings is in bijection with the set of all real numbers in the interval
$[0,1)$. If we're being extra pedantic, we need to forbid an infinite trailing string of all $1$s,
since, e.g., $0.0111\ldots = 0.1000\ldots$ after carrying the $1$s (akin to the fact
that, e.g., $0.0999\ldots = 0.1$). After dealing with this detail, one will have shown that
$\bigl| [0,1]\bigr| \ne |\NN|$.

>>>>

\vskip \medskipamount \hbox to\hsize {\relax \unhcopy \strutbox \cont
{{\tenssbx I. Foundations \kern 3pt}}{2}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 1.\kern .5em Set theory }{3}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 2.\kern .5em Propositional logic}{10}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 3.\kern .5em Predicate logic}{17}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 4.\kern .5em Proofs}{21}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 5.\kern .5em Functions}{28}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 6.\kern .5em Cardinality}{33}}
%\smallskip
%\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 7.\kern .5em Relations}{37}}

@@

We define $g : F\to E$ by
$$g(A) = \cases{A\setminus\{x\}, & if $x\in A$;\cr A\cup \{x\}, & if $x\notin A$.\cr}$$
As before, $\bigl|g(A)\bigr|$ is even, since $A$ is assumed to be a member of $F$ now. Now
for any $A\in E$,
$$\eqalign{
g\bigl(f(A)\bigr) &= \cases{g\bigl(A\setminus\{x\}\bigr), & if $x\in A$;\cr
g\bigl(A\cup \{x\}\bigr), & if $x\notin A$\cr} \cr
&= \cases{\bigl(A\setminus\{x\}\bigr)\cup \{x\}, & if $x\in A$;\cr
\bigl(A\cup \{x\}\bigr)\setminus \{x\} , & if $x\notin A$\cr} \cr
&= \cases{A, & if $x\in A$;\cr
A, & if $x\notin A$\cr} \cr
&= A.\cr
}$$
The proof that $f\bigl(g(A)\bigr) = A$ is similar. Thus $g$ is the inverse of $f$.\slug

@@

\datestamp{26}{ix}
Earlier, we defined the cardinality of a set to be the number of elements it contains. What, then,
is the cardinality of $\NN$? How about $\RR$? You might say $\infty$, but this is not a number
(at least, it's not an element of $\NN$, the way all cardinalities of finite sets are).
So perhaps we should amend our question to the following: {\sl When do infinite sets have the same size?}
Our experience with functions leads us to the answer: {\sl When there exists a bijection between
them}. We shall say that $A$ and $B$ are {\it equipotent} (or {\it equinumerous}, or {\it have
the same cardinality}) if there exists a bijection between $A$ and $B$. In this case
we write $|A| = |B|$.

@@

So we can find subsets of $\NN$ equipotent with it. It turns out we can also find supersets of $\NN$ with
the same property.
\proclaim Theorem \advthm. We have $|\NN| = |\ZZ|$.

@@

Note first that if $n$ is even, then $f(n)\ge 0$,
and if $n$ is odd, then $f(n) < 0$. So if $f(m) = f(n)$ for some $m,n\in \NN$, then $f(m)$ and $f(n)$ must
either both be negative, or both be nonnegative. Either way, $m$ and $n$
are either both even or they are both odd. If $m$ and $n$ are both even, then from $f(m) = f(n)$ we derive
$${m\over 2} = {n\over 2},$$
whence multiplying by $2$ on both sides we see that $m=n$. If $m$ and $n$ are both odd, then
$$-{m+1\over 2} = -{n+1\over 2},$$
so, multiplying by $-2$ and subtracting $1$ from both sides we have $m=n$ in this case as well.

@@

\proof
Enumerate $A = \{a_0, a_1, \ldots\}$ and $B = \{b_0, b_1, \ldots\}$. Now given
$(a_i, b_j)\in A\times B$ for some $i,j\in \NN$, we can let
$$f(a_i, b_j) = 2^i3^j.$$
We show that this defines an injective function. Suppose that $f(a_i, b_j) = f(a_{i'}, b_{j'}).$
Then $2^i3^j = 2^{i'} 3^{j'}$, and by the uniqueness of prime factorisations in the Fundamental
Theorem of Arithmetic, we see that $i = i'$ and $j=j'$.

@@

Let $f : \NN\to A$. We shall show that $f$ is not surjective. For all $m,n\in \NN$, let $a_{m,n}$
be the $n$th bit of $f(m)$. Consider the infinite binary string
$$s = (1-a_{0,0},1-a_{1,1}, 1-a_{2,2}, 1-a_{3,3}, \ldots).$$
This string $s$ cannot equal $f(m)$ for any $m\in \NN$, since given an arbitrary $m$,
the $m$th bit of $f(m)$ is $a_{m,m}$, whereas the $m$th bit of $s$ is $1-a_{m,m}$. Hence there is some
$s\in A$ such that $f(m)\ne s$ for all $m\in \NN$. So $f$ is not surjective.

@@

The proof above was published by Georg Cantor in 1891, and hence is known as {\it Cantor's diagonal argument}.
The technique has since been used to prove many other things.

The set of all infinite binary strings is in bijection with the set of all real numbers in the interval
$[0,1)$. If we're being extra pedantic, we need to forbid an infinite trailing string of all $1$s,
since, e.g., $0.0111\ldots = 0.1000\ldots$ after carrying the $1$s (akin to the fact
that, e.g., $0.0999\ldots = 0.1$). After dealing with this detail, one will have shown that
$\bigl| [0,1]\bigr| \ne |\NN|$.
\bye
